Population structure and phenotypic variation of *Sclerotinia sclerotiorum* from dry bean in the United States

### 2017-08-14 doc/manuscript/manuscript.Rmd
update manuscript up to genetic diversity
I also moved the poppr table to the main text and
futzed with it.

Things I can't address right now:

 - how sclerotia was moved from field to lab
 - whether or not the DS media was in-fact
   modified (I got that from Otto-Hanson)
 - manual scoring of alleles

Things I won't address:

 - Changing years to two-digit years: did you
   forget about Y2K? Four-digit years give more
   information and we have the space.

Several of the isolates used for this study were collected and reported in previous studies and were collected using the same methods [@otto-hanson2011variation].
Broadly, isolates were collected from two sources, white mold screening nurseries (wmn), or producer fields.
The white mold screening nurseries were 5m x 10m in size and maintained without applying fungicides to observe natural incidence of white mold.
The nursery plots were incorporated with a basal dressing of N:P:K = 1:3:2 and side dressing of 0:3:2 during the growing season [@steadman2003identification; @steadman2004identification; @steadman2005identification; @steadman2006identification; @otto2007identification].
Sampling was carried out by collecting diseased tissue in zig-zag transects across field plots. 
Because sampling depended on disease incidence, the number of samples isolated varied from year to year.
Although the nursery locations were the same over sampling years, sampling plots within a location varied for sampling years.

Sclerotia of *S. sclerotiorum* were collected over several years from grower fields and/or white mold screening nurseries in 11 states of the United States of America, Mexico, France, and Australia (Table \@ref(tab:isolate-table)).
The white mold screening nurseries in different regions are indicated by wmn (white mold screening nursery) and all other isolates came from commercial fields (Table \@ref(tab:isolate-table)).

@@

MCG was determined as described previously through co-culturing pairs of 2-day-old isolates 2.5cm apart on Diana Sermons (DS) Medium [@cubeta2001mycelial]. 
Incompatibility of different MCGs resulted in formation of a mycelial barrier of dead tissue (referred to as a barrage line) accompanied by the formation of sclerotia on either side of the barrage line, indicating the limits of the isolates' growth [@otto-hanson2011variation; @kohn1990mycelial; @leslie1993fungal].
Isolates were compared in a pairwise manner for each site and then representatives among sites were compared to determine mycelial compatibility groups by scoring compatible and incompatible interactions [@otto-hanson2011variation].

@@

Aggressiveness of each isolate was assessed using a straw test as carried out in @otto-hanson2011variation that rated necrotic lesion size [@petzoldt1996straw; @teran2006modified].
Briefly, the straw test uses 21-day-old G122 plants as the host in a greenhouse setting. 
Clear drinking straws cut to 2.5cm and sealed were used to place two mycelial plugs of inoculum on the host plant after removing the everything &gt;2.5cm above the fourth node.
Measurements were taken after 8 days using the Modified Petzoldt and Dickson scale where a score of 1--9 where 1 is no disease and 9 is plant death [@petzoldt1996straw; @teran2006modified].

@@

DNA was purified from 100 mg mycelium collected from an actively growing culture of each isolate using a Qiagen DNeasy Plant Mini Kit (Qiagen Inc., Valencia, CA) according to the manufacturer's instructions.

@@

We genotyped *S. sclerotiorum* isolates using 16 microsatellite primer pairs developed previously [@sirjusingh2001characterisation].
PCR was carried out as described using primers labeled with FAM fluorophore, and resulting PCR products were first resolved in a 1.5% agarose gel stained with ethidium bromide to ensure product was within the expected size range.

@@

## Data processing and Analyses {-}

All analyses were performed on this data set unless otherwise mentioned.
All data processing and analyses were performed in R version 3.4.1 [@R] and are openly available and reproducible at https://github.com/everhartlab/sclerotinia-366/.
Of the 16 microsatellite loci genotyped, five included compound repeats, which made it challenging to accurately/confidently bin alleles into fragment sizes expected for each locus based on the described repeat motif. 
<!-- add something about repeat length consistency  -->
Loci with compound repeats were removed for the reported statistics, but were processed and included in data analyses to ensure the integrity of the results with 11 loci as compared to 16.
We assessed the power of our 11 markers by generating a genotype accumulation curve in the R package *poppr* version 2.4.1, looking for evidence of saturation, which would indicate that loci were sufficiently sampled to adequately represent the full set of genotypes [@arnaud2007standardizing; @kamvar2015novel].
To avoid including samples that were potentially collected from the same plant (which increases the probability of collecting the same sample more than once), data were clone-corrected on the region/source/host/year hierarchy for subsequent analyses. The clone-corrected data set included 318 isolates.

@@

A distance-based redundancy analysis (dbRDA) [@legendre1999distance] was performed with the function `capscale()` in the *vegan* package version 2.4-3 [@vegan].
This method uses constrained ordinations on a distance matrix representing the response variable to delineate relative contribution of any number of independent explanatory variables. 
We used this method to delineate the phenotypic (Aggressiveness, MCG), geographic (Region, Host, Location), and temporal (Year) components in influencing genetic composition of the populations.
<!-- RDA is a multivariate linear regression analysis of independent (explanatory: eg. environmental and phenotypic) variables on dependent (non-explanatory or response) variables. -->
The distance matrix we used as the response variable was generated using Bruvo's genetic distance from clone-corrected data on the hierarchy of Region, Location, Host, Year as implemented in *poppr* version 2.4.1, which employed a stepwise mutation model for microsatellite data [@bruvo2004simple; @kamvar2014poppr]. 
Because aggressiveness measures differed with each isolate, aggressiveness first averaged across clone-corrected isolates.
<!-- In our case, a pairwise Bruvo's [@bruvo2004simple] distance matrix was generated from microsatellite data using *poppr* 2.4.1 [@kamvar2014poppr; @kamvar2015novel] and employed it as the dependent matrix for RDA analysis. -->
To identify explanatory variable(s) correlated with genetic variation, a forward-backward selection process was applied with the vegan function `ordistep()` to identify the set of variables that explained more variation than predicted with the global model containing all variables.
An ANOVA was then performed to test for significance of the reduced model and marginal effects using 999 permutations.
The `varpart()` function of vegan was used to determine variation partitioning of explanatory variables.

@@

We used analysis of variance (ANOVA) to assess if aggressiveness could be explained by Region, mycelial compatibility group, or multilocus genotype. 
To minimize complications due to small sample sizes, we chose the top 10 MCGs, representing 56.5% of the isolates collected, the top 10 MLGs representing 26.7% of the isolates, and populations with more than five isolates.
Between-group differences were assessed using Tukey's HSD test with an alpha of 0.05 using the `HSD.test()` function in the package *agricolae* version 1.2.4 [@mendiburu2015agricolae].

@@

We wanted to assess if there was correlation between multilocus genotypes (MLG) and mycelial compatibility groups (MCG). 
This was done using a graph-based approach with community analysis.
We first created a contingency table with MLGs and MCGs and converted it to a directed and weighted edgelist where each edge was a connection from an MCG to an MLG weighted by the number of samples shared in the connection. 
This was then converted to a bipartite graph where top nodes represented MCGs and bottom nodes represented MLGs.
The cluster walktrap community detection algorithm as implemented in the `cluster_walktrap()` function in *igraph* version 1.1.2 was used to define clusters of MLGs and MCGs [@csardi2006igraph; @pons2006computing].
All communities that had fewer than 10 members were then consolidated into one.
These community definitions were then used to assess the average genetic distance (as defined by Bruvo's distance) within members of the community [@bruvo2004simple].

@@

@@

```{r table_ia, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
suppressPackageStartupMessages(library("tidyverse"))
read_csv("../../results/tables/genotype_diversity_table_region.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(
         Pop = col_character(),
         N = col_integer(),
         MLG = col_integer(),
         eMLG = col_double(),
         SE = col_double(),
         H = col_double(),
         G = col_double(),
         lambda = col_double(),
         E.5 = col_double(),
         Hexp = col_double(),
         Ia = col_double(),
         p.Ia = col_double(),
         rbarD = col_double(),
         p.rD = col_double(),
         File = col_character()
      )) %>%
  mutate(Ia = case_when(
    !is.finite(p.Ia) ~ "-",
    p.Ia == 0.001 ~ paste0(round(Ia, 2), "*"),
    p.Ia <= 0.01  ~ paste0(round(Ia, 2), "-"),
    p.Ia <= 0.05  ~ paste0(round(Ia, 2), "~"),
    TRUE          ~ as.character(round(Ia, 2))
  )) %>%
  mutate(rbarD = case_when(
    !is.finite(p.rD) ~ "-",
    p.rD == 0.001 ~ paste0(round(rbarD, 2), "*"),
    p.rD <= 0.01  ~ paste0(round(rbarD, 2), "-"),
    p.rD <= 0.05  ~ paste0(round(rbarD, 2), "~"),
    TRUE          ~ as.character(round(rbarD, 2))
  )) %>%
  mutate_if(is.numeric, round, digits = 2) %>%
  mutate(N = paste(N, MLG, sep = " (") %>% paste0(")") %>% formatC()) %>%
  mutate(eMLG = paste(eMLG, SE, sep = " (") %>%  paste0(")") %>% formatC()) %>%
  rename(`*h*` = Hexp) %>%
  rename(`$E_5$` = E.5) %>%
  rename(`$I_A$` = Ia) %>%
  rename(`$\\lambda$` = lambda) %>%
  rename(`$\\bar{r}_d$` = rbarD) %>%
  select(-File, -MLG, -SE, -p.rD, -p.Ia) %>% 
  huxtable::as_huxtable(add_colnames = TRUE) %>%
  huxtable::set_number_format(huxtable::everywhere, 4:5, 1) %>%
  huxtable::set_align(huxtable::everywhere, -1, "right") %>%
  huxtable::set_col_width(c(1.1, 0.9, 1.3, 0.5, 0.5, 1.2, 0.7, 0.6, 0.6, 0.8)) %>%
  huxtable::print_md(max_width = 80)
```

Table: (\#tab:iatab) Genotypic diversity and Linkage Disequilibrium summary for geographic populations arranged by abundance after clone-correction by a hierarchy of Region/Source/Host/Year. 
Pop = Population, N = number of individuals (number of MLG in parentheses), 
eMLG = expected number of MLGs based on rarefaction at min(10, N) individuals (standard error in parentheses), 
H = Shannon-Weiner Index, G = Stoddardt and Taylor's Index, 
$\lambda$ = Simpson's Index, 
*h* = Nei's 1978 gene diversity, 
$E_5$ = Evenness,
$I_A$ = the index of association, 
$\bar{r}_d$ = standardized index of association. 
P-value symbols: \* = 0.001, - = $\leq 0.01$, ~ = $\leq 0.05$.

>>>>

Several of the isolates used for this study were collected and reported in previous studies and were collected using the same methods [@otto-hanson2011variation].
Broadly, isolates were collected from two sources, white mold screening nurseries (wmn) or producer fields.
The white mold screening nurseries were 5m x 10m in size and maintained without applying fungicides to observe natural incidence of white mold.
The nursery plots were incorporated with a basal dressing of N:P:K = 1:3:2 and side dressing of 0:3:2 during the growing season [@steadman2003identification; @steadman2004identification; @steadman2005identification; @steadman2006identification; @otto2007identification].
Sampling was carried out by collecting diseased tissue in zig-zag transects across field plots. 
Because sampling depended on disease incidence, the number of samples isolated varied from year to year.
Although the nursery locations were the same over sampling years, sampling plots within a location varied for sampling years.

Sclerotia of *S. sclerotiorum* were collected over several years from grower fields and/or wmn in 11 states of the United States of America, Mexico, France, and Australia.
The white mold screening nurseries in different regions are indicated by wmn and all other isolates came from commercial fields (Table \@ref(tab:isolate-table)).

@@

MCG was determined as described previously through co-culturing pairs of 2-day-old isolates 2.5cm apart on Diana Sermons (DS) Medium [@cubeta2001mycelial]. 
Incompatibility of different MCGs resulted in formation of a mycelial barrier of dead tissue (termed a barrage line) accompanied by formation of sclerotia on either side of the barrage line, indicating the limits of the isolates' growth [@otto-hanson2011variation; @kohn1990mycelial; @leslie1993fungal].
Isolates were compared in a pairwise manner for each site and then representatives among sites were compared to determine mycelial compatibility groups by scoring compatible and incompatible interactions [@otto-hanson2011variation].

@@

Aggressiveness of each isolate was assessed using a straw test as described in @otto-hanson2011variation that rated necrotic lesion size [@petzoldt1996straw; @teran2006modified].
Briefly, the straw test uses 21-day-old G122 plants as the host in a greenhouse setting. 
Clear drinking straws cut to 2.5cm and sealed were used to place two mycelial plugs of inoculum on the host plant after removing all plant growth &gt;2.5cm above the fourth node.
Measurements were taken after 8 days using the Modified Petzoldt and Dickson scale where a score of 1--9 where 1 is no disease and 9 is plant death [@petzoldt1996straw; @teran2006modified].

@@

DNA was purified from approximately 100 mg mycelium collected from an actively growing culture of each isolate using a Qiagen DNeasy Plant Mini Kit (Qiagen Inc., Valencia, CA) according to the manufacturer's instructions.

@@

We genotyped *S. sclerotiorum* isolates using 16 microsatellite primer pairs developed previously [@sirjusingh2001characterisation].
PCR was carried out as described using primers labeled with FAM fluorophore, and resulting amplicons were first resolved in a 1.5% agarose gel stained with ethidium bromide to ensure product was within the expected size range prior to capillary electrophoresis.

@@

## Data processing {-}

All analyses were performed on this data set unless otherwise mentioned.
All data processing and analyses were performed in R version 3.4.1 [@R] and are openly available and reproducible at https://github.com/everhartlab/sclerotinia-366/.
Of the 16 microsatellite loci genotyped, five included compound repeats, which made it challenging to accurately/confidently bin alleles into fragment sizes expected for each locus based on the described repeat motif. 
<!-- add something about repeat length consistency  -->
Loci with compound repeats were removed for the reported statistics, but were processed and included in data analyses to ensure the integrity of the results with 11 loci as compared to 16.
We assessed the power of our 11 markers by generating a genotype accumulation curve in the R package *poppr* version 2.4.1, looking for evidence of saturation, which would indicate that loci were sufficiently sampled to adequately represent the full set of genotypes [@arnaud2007standardizing; @kamvar2015novel].
To avoid including isolates potentially collected from the same plant (which increases the probability of collecting sclerotioa from the same point of infection more than once), data were clone-corrected on a hierarchy of region/source/host/year---meaning that one representative of each genotype was retained for each combined regions, source, host, and year---for subsequent analysis.

@@

A distance-based redundancy analysis (dbRDA) [@legendre1999distance] was performed with the function `capscale()` in the *vegan* package version 2.4-3 [@vegan].
This method uses constrained ordinations on a distance matrix representing the response variable to delineate relative contribution of any number of independent explanatory variables. 
We used this method to delineate the phenotypic (Aggressiveness, Mycelial Compatibility Group (MCG)), geographic (Region, Host, Location), and temporal (Year) components in predicting genetic composition of the populations.
<!-- RDA is a multivariate linear regression analysis of independent (explanatory: eg. environmental and phenotypic) variables on dependent (non-explanatory or response) variables. -->
The distance matrix we used as the response variable was generated using Bruvo's genetic distance from clone-corrected data (procedure described above) as implemented in *poppr* version 2.4.1, which employed a stepwise mutation model for microsatellite data [@bruvo2004simple; @kamvar2014poppr]. 
Because aggressiveness measures differed with each isolate, aggressiveness first averaged across clone-corrected isolates.
<!-- In our case, a pairwise Bruvo's [@bruvo2004simple] distance matrix was generated from microsatellite data using *poppr* 2.4.1 [@kamvar2014poppr; @kamvar2015novel] and employed it as the dependent matrix for RDA analysis. -->
To identify explanatory variable(s) correlated with genetic variation, a forward-backward selection process was applied with the vegan function `ordistep()` to identify the set of variables that explained more variation than predicted with the global model containing all variables.
An ANOVA was then performed to test for significance of the reduced model and marginal effects using 999 permutations.
The `varpart()` function of vegan was used to determine variation partitioning of explanatory variables.

@@

We used analysis of variance (ANOVA) to assess if aggressiveness could be explained by Region, MCG, or multilocus genotype (MLG). 
To minimize complications due to small sample sizes, we chose the top 10 MCGs, representing 56.5% of the isolates collected, the top 10 MLGs representing 26.7% of the isolates, and populations with more than five isolates.
Between-group differences were assessed using Tukey's HSD test with an alpha of 0.05 using the `HSD.test()` function in the package *agricolae* version 1.2.4 [@mendiburu2015agricolae].

@@

We wanted to assess if there was correlation between MLGs and MCGs. 
In simple terms, this was done using a network-based approach where both MLGs and MCGs were considered nodes and the number of isolates in which they were found together was the strength of the connection between an MLG and and MCG node.
The network-based approach allowed us to assess how "clumpy" the relationships between MLGs and MCGs were.
In more specific terms, we first created a contingency table with MLGs and MCGs and converted it to a directed and weighted edgelist where each edge was a connection from an MCG to an MLG weighted by the number of samples shared in the connection. 
This was then converted to a bipartite graph where top nodes represented MCGs and bottom nodes represented MLGs.
The cluster walktrap community detection algorithm as implemented in the `cluster_walktrap()` function in *igraph* version 1.1.2 was used to define clusters of MLGs and MCGs [@csardi2006igraph; @pons2006computing].
All communities that had fewer than 10 members were then consolidated into one.
Community definitions were used to assess the average genetic distance (as defined by Bruvo's distance) within members of the community [@bruvo2004simple].

@@

```{r table_ia, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
suppressPackageStartupMessages(library("tidyverse"))
read_csv("../../results/tables/genotype_diversity_table_region.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(
         Pop = col_character(),
         N = col_integer(),
         MLG = col_integer(),
         eMLG = col_double(),
         SE = col_double(),
         H = col_double(),
         G = col_double(),
         lambda = col_double(),
         E.5 = col_double(),
         Hexp = col_double(),
         Ia = col_double(),
         p.Ia = col_double(),
         rbarD = col_double(),
         p.rD = col_double(),
         File = col_character()
      )) %>%
  # mutate(Ia = case_when(
  #   !is.finite(p.Ia) ~ "-",
  #   p.Ia == 0.001 ~ paste0(round(Ia, 2), "*"),
  #   p.Ia <= 0.01  ~ paste0(round(Ia, 2), "-"),
  #   p.Ia <= 0.05  ~ paste0(round(Ia, 2), "~"),
  #   TRUE          ~ as.character(round(Ia, 2))
  # )) %>%
  mutate(rbarD = case_when(
    !is.finite(p.rD) ~ "-",
    p.rD == 0.001 ~ paste0(round(rbarD, 2), "*"),
  #   p.rD <= 0.01  ~ paste0(round(rbarD, 2), "-"),
  #   p.rD <= 0.05  ~ paste0(round(rbarD, 2), "~"),
    TRUE          ~ as.character(round(rbarD, 2))
  )) %>%
  mutate_if(is.numeric, round, digits = 2) %>%
  mutate(N = paste(N, MLG, sep = " (") %>% paste0(")") %>% formatC()) %>%
  mutate(eMLG = paste(eMLG, SE, sep = " (") %>%  paste0(")") %>% formatC()) %>%
  rename(`*h*` = Hexp) %>%
  rename(`$E_5$` = E.5) %>%
  # rename(`$I_A$` = Ia) %>%
  rename(`$\\lambda$` = lambda) %>%
  rename(`$\\bar{r}_d$` = rbarD) %>%
  select(-File, -MLG, -SE, -p.Ia, -p.rD, -Ia) %>% 
  huxtable::as_huxtable(add_colnames = TRUE) %>%
  huxtable::set_number_format(huxtable::everywhere, 4:5, 1) %>%
  huxtable::set_align(huxtable::everywhere, -1, "right") %>%
  huxtable::set_col_width(c(1.1, 0.9, 1.3, 0.5, 0.5, 1.1, 0.7, 0.6, 0.7)) %>%
  huxtable::print_md(max_width = 80)
```

Table: (\#tab:iatab) Genotypic diversity and Linkage Disequilibrium summary for geographic populations arranged by abundance after clone-correction by a hierarchy of Region/Source/Host/Year. 
Pop = Population, N = number of individuals (number of MLG in parentheses), 
eMLG = expected number of MLGs based on rarefaction at min(10, N) individuals (standard error in parentheses), 
H = Shannon-Weiner Index, G = Stoddardt and Taylor's Index, 
$\lambda$ = Simpson's Index, 
*h* = Nei's 1978 gene diversity, 
$E_5$ = Evenness,
$\bar{r}_d$ = standardized index of association.
An asterix indicates a significant value of $\bar{r}_d$ after 999 permutations, $P \leq 0.001$.

@@
